A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-23 tex/BiSig.tex
[ update ] Fine-tune the paper outline

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}
\subsection{Type Checking via Bidirectional Typing}

\section{A Definition of Simple Type Theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple) type signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
  An element $i : I$ is meant to be the \emph{$i$-th operation} $\tyop{i}$ and $\mathit{ar}(i)$ the \emph{number} of arguments of $\tyop{i}$.
\end{definition}
\LTnote{Define $\mathsf{Type}$}
\begin{figure}
  \begin{subfigure}{.5\textwidth}
    \begin{mathpar}
      \boxed{\Theta\vdash A} \\
      \inferrule{\Theta \ni X_i}{\Theta \vdash X_i} \and
      \inferrule{\Theta \vdash A_1 \\ \cdots \\ \Theta \vdash A_n}{\Theta \vdash \tyop{i}(A_1, \ldots, A_n)}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}{.4\textwidth}
    \begin{mathpar}
      \boxed{\Theta \vdash \Gamma} \\
      \inferrule{ }{\Theta \vdash \emptyset}\and
      \inferrule{\Theta \vdash A \\ \Theta\vdash \Gamma}{\Theta \vdash \Gamma, A}
    \end{mathpar}
  \end{subfigure}
  \caption{Simple types and contexts}
\end{figure}

@@

\subsection{Binding Signatures and Typed Terms}
\begin{definition}
  For a type signature, a \emph{term signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
  \begin{enumerate}
    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target and
    \item $\left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]$ is a finite sequence of pairs of a context $\Delta_{k, j} : \mathsf{Cxt}(\Xi_j)$ for types of its binding variables and its type $A_{k, j} : \mathsf{Type}(\Xi_j)$.
  \end{enumerate}
\end{definition}
\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash t : A} \\
    \inferrule{\Gamma \ni (x : A)}{\Theta \mid \Gamma \vdash \term{x} : A}\;(\text{var})
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
    \quad \text{where $\rho \colon \Xi_j \to \Theta$}
  \end{mathpar}

@@

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \\
    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j}} \simsub{A_{n_j ,j}}{\rho}}
    {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(\vec{x}_1 .\, t_1, \ldots, \vec{x}_{n_j}.\, t_{n_j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
  \end{mathpar}
  \caption{Bidirectional typing rules}
\end{figure}

@@

\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
\subsubsection{Intrinsically Typed Terms}

@@

\begin{figure}
  \caption{Bidirectional raw terms}
\end{figure}
\subsubsection{Extrinsically Typed Terms}

@@

\section{Formal Implementation and Examples}

@@

\section{Concluding Remarks and Future Work}
\begin{acks}
We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.

>>>>

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}
\subsection{Type Checking via Bidirectional Typing}
\subsection{Formalisation in \Agda}

\section{A Definition of Simple Type Theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple type) signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
  An element $i : I$ is meant to be the \emph{$i$-th operation} $\tyop{i}$ and $\mathit{ar}(i)$ the \emph{number} of arguments of $\tyop{i}$.
\end{definition}
\LTnote{Define $\mathsf{Type}$}
\begin{figure}
  \begin{subfigure}{.5\textwidth}
    \begin{mathpar}
      \boxed{\Theta\vdash A} \\
      \inferrule{\Theta \ni X_i}{\Theta \vdash X_i} \and
      \inferrule{\Theta \vdash A_1 \\ \cdots \\ \Theta \vdash A_n}{\Theta \vdash \tyop{i}(A_1, \ldots, A_n)}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}{.4\textwidth}
    \begin{mathpar}
      \boxed{\Theta \vdash \Gamma} \\
      \inferrule{ }{\Theta \vdash \emptyset}\and
      \inferrule{\Theta \vdash A \\ \Theta\vdash \Gamma}{\Theta \vdash \Gamma, A}
    \end{mathpar}
  \end{subfigure}
  \caption{Simple types and contexts}
\end{figure}

@@

\subsection{Binding Signatures and Intrinsically Typed Terms}
\begin{definition}
  For a type signature, a \emph{binding signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
  \begin{enumerate}
    \item $\Xi_j : \mathbb{N}$ is the number of variables, 
    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target,  and
    \item $\left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]$ is a finite sequence of pairs of a context $\Delta_{k, j} : \mathsf{Cxt}(\Xi_j)$ for types of its binding variables and its type $A_{k, j} : \mathsf{Type}(\Xi_j)$.
  \end{enumerate}
\end{definition}
\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash t : A} 
    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} \\
    \inferrule{\Gamma \ni (x : A)}{\Theta \mid \Gamma \vdash \term{x} : A}\;(\text{var})
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j}(\rho; t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
    \quad \text{where $\rho \colon \Xi_j \to \Theta$}
  \end{mathpar}

@@

@@

\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}

\subsubsection{Intrinsically Typed Terms}

@@

\begin{figure}
  \caption{Bidirectional raw terms}
\end{figure}

\subsubsection{Extrinsically Typed Terms}

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \quad
    \text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$}\\
    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j, j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j, j}} \simsub{A_{n_j ,j}}{\rho}}
    {\Theta \mid \Gamma \vdash \term{\op{j}(\rho; \vec{x}_{1, j} .\, t_{1, j}, \ldots, \vec{x}_{n_j, j}.\, t_{n_j, j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
  \end{mathpar}
  \caption{Bidirectional typing rules}
\end{figure}

@@

\section{Formal Implementation with Examples}

@@

\section{Future Work}
\begin{acks}
We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
